Iterative Unfolding Optimization with the Mean Squared Error Metric

### 2019-09-28 NOvA/numu/mean_squared_error_metric/mse_metric.tex
Work in progress.

\begin{frame}{Recap}
  \begin{itemize}
    \item The standard metric in the ND group used by all analyses requiring unfolding is the average global correlation coefficient\footnote[frame]{Stefan Schmitt, ``Data Unfolding Methods in High Energy Physics''},
      \begin{equation} \label{eq:1}
        \rho_{avg}=\frac{1}{M}\sum_{j=1}^{M}\sqrt{1-\frac{1}{\pmb{V}_{jj}(\pmb{V}^{-1})_{jj}}}
      \end{equation}
    , where $M$ is the number of bins and $\pmb{V}$ is the covariance matrix in true space inferred by the unfolding algorithm.
    \item For analyses with tens of bins, this is a convenient metric. However, for an analysis with thousands of bins, this metric turned out to be infeasible.
  \end{itemize}
\end{frame}

\begin{frame}{Infeasibility of Average Global Correlation Coefficient for Many-Bin Analyses}
  \begin{columns}
    \begin{column}{.65\textwidth}
    Inverting a covariance matrix this large turns out to be very tricky.
    \begin{itemize}
      \item The covariance matrices all have astronomical \textcolor{red}{condition numbers} (i.e., ill-conditioned or nearly singular).
      \item Numerical inversion is still possible but subject to an arbitrary, small cut-off number, or tolerance, brought into play by SVD.
      \item Forcefully getting the calculation through results in numerical instability, such as negative values in square root in Eq.~\ref{eq:1}. Removing unphysical values, results are shown to the right. No clear minimum is observed.
    \end{itemize}
    \end{column}
    \begin{column}{.35\textwidth}
      \includegraphics[width=\textwidth]{figures/cov_mat_iter20.png} \\
      \includegraphics[width=\textwidth]{figures/avg_rho.eps}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}{Reference}
  \begin{columns}
    \begin{column}{.5\textwidth}
      Most of the contents in this document are taken from this textbook, especially Chapter 11 dedicated to unfolding.
    \end{column}
    \begin{column}{.5\textwidth}
      \centering
      \includegraphics[height=\textheight]{figures/cowan_front_cover.jpeg}
    \end{column}
  \end{columns}
\end{frame}

>>>>

%================================================================================
\begin{frame}{Recap}
  \begin{itemize}
    \item The standard metric in the ND group used by all analyses requiring unfolding is the average global correlation coefficient\footnote[frame]{Stefan Schmitt, ``Data Unfolding Methods in High Energy Physics''},
      \begin{equation} \label{eq:1}
        \rho_{avg}=\frac{1}{M}\sum_{j=1}^{M}\sqrt{1-\frac{1}{\pmb{V}_{jj}(\pmb{V}^{-1})_{jj}}}
      \end{equation}
    , where $M$ is the number of bins and $\pmb{V}$ is the covariance matrix in true space inferred by the unfolding algorithm.
    \item For analyses with tens of bins, this is a convenient metric. However, for an analysis with thousands of bins, this metric turned out to be infeasible.
  \end{itemize}
\end{frame}

%================================================================================
\begin{frame}{Infeasibility of Average Global Correlation Coefficient for Many-Bin Analyses}
  \begin{columns}
    \begin{column}{.65\textwidth}
    Inverting a covariance matrix this large turns out to be very tricky.
    \begin{itemize}
      \item The covariance matrices all have astronomical \textcolor{red}{condition numbers} (i.e., ill-conditioned or nearly singular).
      \item Numerical inversion is still possible but subject to an arbitrary, small cut-off number, or tolerance, brought into play by SVD.
      \item Forcefully getting the calculation through results in numerical instability, such as negative values in square root in Eq.~\ref{eq:1}. Removing unphysical values, results are shown to the right. No clear minimum is observed.
    \end{itemize}
    \end{column}
    \begin{column}{.35\textwidth}
      \includegraphics[width=\textwidth]{figures/cov_mat_iter20.png} \\
      \includegraphics[width=\textwidth]{figures/avg_rho.eps}
    \end{column}
  \end{columns}
\end{frame}

%================================================================================
\begin{frame}{A Simple Alternative Metric: Mean Squared Error}
	Suppose $\theta$ is the a true parameter to be estimated, and $\hat{\theta}$ is an estimator of the parameter.
	The mean squared error (MSE) can be decomposed into a sum of variance and bias squared.
	\begin{eqnarray} \label{eq:2}
	MSE &=& E[(\hat{\theta}-\theta)^2]\nonumber\\
	&=& E[\hat{\theta}^2-2\hat{\theta}\theta+\theta^2] = E[\hat{\theta}^2] -2\theta E[\hat{\theta}]+\theta^2 \nonumber\\
	&=& (E[\hat{\theta}^2]-E[\hat{\theta}]^2)+(E[\hat{\theta}]^2-2\theta E[\hat{\theta}]+\theta^2) \nonumber\\
	&=& V[\hat{\theta}]+b^2
	\end{eqnarray}
	, where $b=E[\hat{\theta}]-\theta$ is the bias of the estimator.\\
	Equation~\ref{eq:2} is called bias-variance decomposition.
	\vfill
	This is a very common metric in statistics and machine learning as well.
\end{frame}

%================================================================================
\begin{frame}{Reference}
  \begin{columns}
    \begin{column}{.5\textwidth}
      Most of the contents in this document are taken from this textbook, especially Chapter 11 dedicated to unfolding.
    \end{column}
    \begin{column}{.5\textwidth}
      \centering
      \includegraphics[height=\textheight]{figures/cowan_front_cover.jpeg}
    \end{column}
  \end{columns}
\end{frame}
